---
title: "Data Wrangling - Part II"
subtitle: "StatPREP R Workshops"
author: "Quintin Burnett"
output: html_document
---

```{r setup, include=FALSE}
# Don't forget to load your libraries
library(tidyverse)
library(ggplot2movies)
library(ggformula)
```

```{r}
# This loads the data from the ggplot2movies package
data(movies)
```

# Deriving information

So far, you have learned "verbs" for isolating data, `select`, `filter`, `arrange`. Now, we will learn verbs for deriving information, `summarize`, and `mutate`. There is an additional verb, `group_by`, which is useful in conjunction with these verbs, but not as much on its own. 

## summarize

Summarize is a many-to-one or many-to-few operation. It summarizes the dataset down to fewer numbers. 

Let's say we want to know the average budget of the movies in the `movies` dataset. The way to compute that is with summarize,

```{r}
movies %>%
  summarize(mean(budget))
```

Unfortunately, there are many missing (or Not Available) values in `budget`, so we need to remove those in our pipeline.

```{r}
movies %>%
  drop_na(budget) %>%
  summarize(mean(budget))
```

Notice that this pipeline does not change the `movies` dataset in our environment. The NA values are still present in the data, and the entire dataset is just how it started.  We just `summarize`d the data down to a single number in our Console or RMarkdown document. 

If we want, we can add additional statistics to our `summarize` command,

```{r}
movies %>%
  drop_na(budget) %>%
  summarize(mean(budget), max(budget))
```

We can also name the statistics we compute, so we could refer to them later.

```{r}
movies %>%
  drop_na(budget) %>%
  summarize(avg_budget = mean(budget), max_budget = max(budget))
```

### Question 1

Now, you try it! 

- What year was the earliest movie in the dataset released? 
- What is the average length of movies?
- What is the standard deviation of length?

You may want to look at the "data transformation with dplyr" cheatsheet to review some possible summary functions. 

```{r eval=FALSE}
movies %>%
  summarize(
    earliest_year = min(year, na.rm = TRUE),
    avg_length = mean(length, na.rm = TRUE),
    sd_length = sd(length, na.rm = TRUE)
  )
```

The earliest movie in the dataset was released in the year 1893. The average length of movies is approximately 82.34 minutes. The standard deviation of the length of the movies is around 44.35 minutes.


## Question 2

The `group_by` function groups cases by common values. 

This is useful if want to compute summary statistics by group. For example, we may be interested in whether action movies are longer than non-action movies. 

We already know how to compute the average length of movies overall. What we need to add to our pipeline is a `group_by`

```{r eval=FALSE}
movies %>%
  group_by(Action) %>%  
  summarize(
    avg_length = mean(length, na.rm = TRUE),
    count = n()                               
  )
```

Which movies are longer, action movies or non-action movies? 

Action movies are generally longer than non-action movies, with an average length of 98.99 minutes compared to 80.89 minutes.

### Question 3

Investigate the movies classified as Short. Are they actually shorter than other movies? 

- What is the average length? 
- What is the minimum length?

```{r eval=FALSE}
short_movies <- movies %>%
  group_by(Short) %>%
  summarize(
    avg_length = mean(length, na.rm = TRUE),
    min_length = min(length, na.rm = TRUE)
  )
short_movies
```

Movies that are "Short" are indeed much shorter than other movies, with an average length of only 13.97 minutes compared to 95.45 minutes for non-short movies. Both have their minimum length at 1.

## mutate

The `mutate` function creates new variables (columns) in the dataset. Typically, we use it to make new variables from old ones. 

For example, we may be interested in the length of movies in hours, rather than in minutes. 

```{r}
movies <- movies %>%
  mutate(length_hrs = length / 60)
```

Let's read this code carefully. 

- what does the `<-` do?
- why are we writing `/60`?
- what does the `=` do?
- why did we use the name `length_hrs`?

### Question 4

The variables `r1`, `r2`, ..., `r10` give you the percent of people who rated that movie a 1, 2, ..., 10. We would like to know the percent of people who rated the movie a 4 or less (i.e., who rated it a 1, 2, 3, or 4). Create a new variable called `per_low` that computes that number.

```{r eval=FALSE}
movies %>%
  mutate(per_low = (r1 + r2 + r3 + r4) / 100 * 100) %>%
  select(title, year, per_low) %>%
  head(10)
```

My code adds up all the percentages of people who gave a rating of 1 to 4.

### Question 5

Now that we have a bunch of wrangling skills, let's put them together. 

- pull out the movies that had a high percent of low ratings (Hint: what command will you use to do this?)
- pull out movies that had a high percent of low ratings *and* more than 5 votes. 
- how many of these bad movies were there?

To accomplish that last task, we need a specialty function, `n`. Here's how it works

```{r}
bad_movies_average <- movies %>%
  filter(rating < 5 & votes > 5) %>%
  select(title, rating, votes) %>%
  summarize(
    average_rating = mean(rating, na.rm = TRUE),
    average_votes = mean(votes, na.rm = TRUE),
    bad_movies_count = n()
  )

bad_movies_ratings <- movies %>%
  mutate(per_low = r1 + r2 + r3 + r4) %>%
  filter(per_low > 50, votes > 5) %>%
  summarize(
    bad_movies_count = n(),
    average_per_low = mean(per_low, na.rm = TRUE)
  )

bad_movies_average
bad_movies_ratings

```

To paraphrase my code I took the percentages of people who rated movies 4 and below which is r1 - r4. r5 - r10 are high ratings so they are excluded. I filtered it so that the votes as anything lower than 5 and above 50%. This gave me 4753 movies. I have another piece of code that uses the rating column to show the average ratings of the movies that are bad. Under the average it shows the average movies that are low rated are 13518 from the average rating being 1-10. My code is a rating under 5. 

- what is the average length of movies in the dataset, in hours? 
- can you find a second way to compute the average length, by flipping the order of commands (i.e., `summarize` first and then `mutate`)?

```{r}
average_length_hours <- movies %>%
  summarize(average_length = mean(length, na.rm = TRUE)) %>%
  mutate(average_length_hours_summarize = average_length / 60)

average_length_hours_flipped <- movies %>%
  mutate(length_hours = length / 60) %>%
  summarize(average_length_hours_mutate = mean(length_hours, na.rm = TRUE))

average_length_hours
average_length_hours_flipped

```

The average length of the movies is 1.37 hours.
